---
title: "The Effect of White Nose Syndrome on Little Brown Bats"
author: "Tori Duckworth"
date: "11/12/2024"
output: html_document
---

```{r include = FALSE}
# This chunk just sets up some styling (eg: default size of our images)
knitr::opts_chunk$set(
  collapse = TRUE, 
  warning = FALSE,
  message = FALSE,
  fig.height = 2.75, 
  fig.width = 4.25,
  fig.align = 'center')
```

## ReadME: 
In this project I used data from both the [White-Nose Syndrome Response Team] (https://whitenosesyndrome.org/) as well as a data set from [United States Geological Survery](https://www.sciencebase.gov/catalog/item/620e6f3bd34e6c7e83baa603) (USGS) which detailed the status and trends of North American bats in the summer. The first data set spanned from 2006 to 2024 and detailed the first known apperance of either Pd fungus or white nose syndrome (WNS) for each county. While the second data set only spanned from 2010-2019 and gave the probability of seeing a certain bat (in this case Little Brown Bat) in a state for each year. By wrangling and combining this data I was able to see the effect of WNS on the Little Brown Bat population from 2010-2019. 

## MetaData: 
- Regional_occupancy_trends: Name of original data set with probability of occupancy of bat species by state. 
- Regional_occupancy_trends_MYLU: data set cleaned to contain just probability of occupancy of Little Brown Bats (MYLU)
  - Species: The species of bat, in this case filtered to only MYLU 
  - region: the state 
  - quantity: type of analysis 
    - psi_bar_t: looks at actual probability 
    - delta_bar_t: looks at the change in probability
  - year: the year of the prediction/observation 
  - mean: the mean probability of bats in that region
  - lower: the lower estimation of the probability of bats in that region 
  - upper: the upper estimation of the probability of bats in that region 
- WNS_Map_Data: Name of original data set that states the year and county and appearance of either Pd fungus or WNS. 
- WNS_Map_Data_Simple1: Cleaned map dataset that has has a new column year, filters just to the US and renamed state-prov to region. 
- WNS_Map_Data_Simple2: Further cleaned map dataset to only include, determination, year and region. It gets rid of any years that have repeats, so the unit of observation is year. It upscaled from counties to states. 
  - determination: the status the region 
    - Pd Presumed: it is highly likely the Pd fungus is in the area
    - Pd Positive: Pd fungus was identified in the area, but bats do NOT show symptoms 
    - WNS suspect: it is highly likely that bats have been infected
    - WNS Positive: Bats have been spotted and identified with the disease. 
  - year: the year of the observation 
  - region: the state of the observation 
- Mixed_rows: A data frame that has determination, year and region meant to fill in blank years in WNS_Map_Data_Simple2. It was eventually binded to WNS_Map_Data_Simple2. 
- WNS_Occupancy_Binded: A data set that combines the regional occupancy trends and WNS classification by year. If a region was identified as having a classification, that classification was extended till the end of the observation period for the region. If it did not have a classification it was labeled as no presence. 

## Introduction:  
Little Brown Bats (MYLU) were once the most abundant bat species in the United States ranging all across North America. However they are current endangered and even that status is under review. They only live 6-7 years and have one offspring per year, meaning that they recover slowly from any population decline. Furthermore, They are insectivores and opportunistic feeders which help with pest and disease control (Clare et al., 2011), making them vital ecological players for humans. 

White Nose Syndrome, which is caused by the Pseudogymoascus destructans fungus affect hibernating bats by attacking their exposed skin. It irritates them and wakes them during winter months (Pettit & O'Keefe, 2017) causing them to use up valuable fat reserves for energy (Reeder et al., 2012). As a result the bat leaves the cave searching for food and unable to find any eventually starves or freezes to death. The disease has been present in North America since 2006 and is suspected to have traveled from Europe on human clothes (Frick et al., 2016). Once it infected bats it has primarily been spread between them. 

The importance of bats to human economy and health can not be understated. The decline in bats could cause increase in infections diseases as well as crop insects and pests. This could raise spending on pesticides and crop damage as well as healthcare by billions. If these bats are not saved or recover the consequences could be disasterous. 

## Research Questions: 
1. How is the Little Brown Bats population changing? 
2. Is White Nose Syndrome spreading across the US? 
3. How has White Nose syndrome affected the little brown bat populations over time? 

## Hypotheses: 
1. The Little Brown Bat population was and is decreasing, especially considering that they are now endangered and that status is currently under review. 
2. Yes, White Nose Syndrome is spreading across the US. As an invasive species the animals here have no defence against it, so it should easily spread. 
3. White Nose Syndrome is the cause of the species decline and a good predictor that populations in an area will decline. The Pd fungus when it causes White Nose syndrome is incredibly deadly to the bats as it disrupts their hibernation and they have a low birth/replacement rate.   

## Code to set up data 

#### Libraries 
```{r}
library(auk)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(readr)
library(sf)
library(rnaturalearth)
library(leaflet)
library(maps)
library(ggthemes)
library(tidyverse)
library(scales)
library(GLMMadaptive)
library(plotly)
library(rjson)
library(usmap)
```

#### Import the Data 
```{r}
Regional_occupancy_trends <- read.csv("regional_occupancy_trends_species_year.csv")
WNS_Map_Data <- read.csv("wns_county_status (1)/wns_county_status.csv")
```

#### Filtering to just little brown bats 
```{r}
Regional_occupancy_trends_MYLU <- Regional_occupancy_trends %>%
  filter(species == "MYLU") %>%
  filter(region != "Range") %>%
  mutate(region = str_to_lower(region)) %>%
  mutate(year = as.character(year)) 
```

#### Cleaning WNS data for Maping 
```{r}
WNS_Map_Data_Simple1 <- WNS_Map_Data %>%
  mutate(year = str_extract(status_season, "^[0-9]+")) %>%
  mutate(stateprov = str_to_lower(stateprov)) %>%
  mutate(region = stateprov) %>%
  filter(country != "CAN")
```

#### Cleaning Map data from county to state for each year
```{r}
Mixed_rows <- data.frame(determination = c("WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "Pd Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "Pd Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "Pd Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive", "WNS Positive"), 
                         year = c("2016", "2015", "2014", "2013", "2014", "2023", "2013", "2015", "2017", "2012", "2013", "2010", "2008", "2014", "2017", "2016", "2021", "2015", "2013", "2011", "2012", "2020", "2021", "2022", "2016", "2008", "2018", "2013", "2012", "2020", "2012", "2017", "2016", "2009", "2011", "2017", "2013", "2020", "2023", "2010", "2009", "2023", "2011", "2016", "2015", "2014", "2022", "2021"), 
                         region = c("alabama", "alabama", "alabama", "alabama", "arkansas", "california", "georgia", "iowa", "kansas", "kentucky", "kentucky", "maryland", "massachusetts", "michigan", "minnesota", "minnesota", "mississippi", "missouri", "missouri", "missouri", "missouri", "montana", "montana", "montana", "nebraska", "new york", "north carolina", "north carolina", "north carolina", "north dakota", "ohio", "oklahoma", "oklahoma", "pennsylvania", "pennsylvania", "south dakota", "tennessee", "texas", "texas", "virginia", "virginia", "washington", "west virginia", "wisconsin", "wisconsin", "wisconsin", "wyoming", "wyoming"))

WNS_Map_Data_Simple2 <- WNS_Map_Data_Simple1 %>%
  reframe(determination, year, region) %>%
  distinct() %>%
  arrange(region)
  
WNS_Map_Data_Simple2 <- WNS_Map_Data_Simple2[-c(1:8, 13:15, 20, 22, 32, 35, 38:40, 51, 53:55, 57, 61:65, 70, 72, 76, 78:82, 85:89, 94, 95, 101, 103:116, 119:121, 131, 132, 137, 138, 140, 143, 144, 145, 148, 149, 151:153, 156:159, 163, 164, 168, 169, 175, 176, 179, 185, 187, 189, 192, 193, 201:204, 206:208, 217, 219, 220:222, 224:227, 229, 231:233), ]  

WNS_Map_Data_Simple2 <- rbind(WNS_Map_Data_Simple2, Mixed_rows)
```

#### Binding together regional data & WNS data 
```{r}
WNS_Occupancy_Binded <- Regional_occupancy_trends_MYLU %>%
  full_join(WNS_Map_Data_Simple2) %>%
  arrange(region, year)  %>%
  filter(!is.na(year)) %>%
  filter(!is.na(quantity)) 
```

#### Filling in blank years 
```{r}
#important to note that first data set only recorded when it was that condition was FIRST noticed, it is then presumed the condition remained in the rest of the years. Which is why it can be edited in the following way. 
WNS_Occupancy_Binded[c(1, 20:38, 58:77, 115:129, 134:152, 172, 191:209, 210:228, 229:231, 248:266, 267:269, 305:311, 324:336, 362:380, 438:440, 457:461, 476:486, 495:499, 514, 533:549, 552:562, 571:589, 628:646, 647:665, 704:720, 761:769, 799:817, 875:883, 913:915, 932:944, 970:980, 989:1007, 1008:1026, 1046:1054, 1084:1088, 1103:1121), 9] = "No Presence"
WNS_Occupancy_Binded[c(132, 133), 9] = "Pd Presumed"
WNS_Occupancy_Binded[c(4, 5, 502:505, 508:511, 983, 984), 9] = "Pd Positive"
WNS_Occupancy_Binded[c(14:19, 88:95, 154:171, 175:190, 236:247, 272, 273, 278:285, 289:304, 316:323, 339:342, 352:361, 384:399, 420:437, 443:448, 451:454, 468:475, 491:494, 521:532, 565:570, 692:703, 749:760, 776:779, 821:836, 886:893, 920:931, 947:950,  960:969, 1032:1045, 1057:1060, 1068:1073, 1076:1083, 1097:1102), 9] = "WNS Positive"
WNS_Occupancy_Binded[c(80, 81, 609:627, 667, 668, 671:684), 9] = "WNS Suspect"
```

#### Removing places with no determination (Canada)
```{r}
WNS_Occupancy_Binded <- WNS_Occupancy_Binded %>%
  filter(!is.na(determination))
```

## Plots & Code: 

#### Probability of seeing little brown bats over time 
```{r}
Regional_occupancy_trends_MYLU %>%
  filter(quantity == "psi_bar_t") %>%
  ggplot(aes(x = year, y = mean)) + 
  geom_boxplot() + 
  labs(x = "Year", y = "Mean", title = "Mean Probability of Seeing Little Brown Bats over time") + 
  theme_classic()
```
There is a clear decrease in the mean probability of seeing little brown bats over time. Furthermore there is more variability in 2019 because some states have been hit by WNS while others have yet to be hit. 

#### Count of WNS Over time (Counts per year)
```{r}
WNS_Occupancy_Binded %>%
  filter(quantity == "psi_bar_t") %>%
  ggplot(aes(x = year, fill = determination)) + 
  geom_bar() + 
  theme_classic() + 
  labs(x = "Year", y = "Total States", fill = "Status4", title = "Classification of States Status on WNS per year")
```
There is a clear steady increase in the number of states that have WNS. Over half the states are known to have WNS by 2015, and at least 35 by 2019. 

#### Map of probability of WNS in 2010 
```{r}
states_map <- map_data("state")
WNS_states_data <- inner_join(WNS_Map_Data_Simple1, states_map, by = "region")

WNS_Occupancy_Binded  %>%
  filter(year == 2010) %>%
  mutate(region = str_to_lower(region)) %>%
  ggplot(aes(map_id = region,  fill = determination)) + 
  geom_map(map = states_map, color = "black") + 
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map() + 
  labs(title = "Map of WNS state status in 2010", fill = "Status") + 
  scale_fill_manual(values = c("yellow", "darkorchid4", "pink")) 
```
Only the East Coast is really effected by white nose syndrome in 2010.  

#### Map of probability of WNS in 2019
```{r}
WNS_Occupancy_Binded %>%
  filter(year == 2019) %>%
  mutate(region = str_to_lower(region)) %>%
  ggplot(aes(map_id = region, fill = determination)) + 
  geom_map(map = states_map, color = "black") + 
  expand_limits(x = states_map$long, y = states_map$lat) + 
  scale_fill_manual(values = c("yellow", "chartreuse3", "cadetblue", "darkorchid4", "pink")) + 
  theme_map() + 
  labs(fill = "Status", title = "Map of WNS Status in 2019")
```
By 2019 WNS has spread beyond the East Coast and has reach the midwest. Comparing it to the previous map we can see a clear spread from east to west coast across the Unites States. 

#### Map of probability of Bats in 2010 (MYLU)
```{r}
states_map <- map_data("state")

Regional_occupancy_trends_MYLU %>%
  filter(year == "2010") %>%
  filter(region != c("Alberta", "British Columbia", "Manitoba", "New Brunswick", "Nova Scotia", "Ontario", "Prince Edward Island", "Quebec", "Saskatchewan", "Yukon")) %>% 
  mutate(region = str_to_lower(region)) %>%
  ggplot(aes(map_id = region, fill = mean)) + 
  geom_map(map = states_map, color = "black") + 
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map() + 
  scale_fill_viridis_c() + 
  labs(fill = "Mean probability", title = "Probability of occupancy for Little Brown bat by state in 2010")
```
In this map you can see that the probably of seeing bats in east coast states is in fact relatively high, despite the fact that WNS is confirme din those states. 

#### Map of probability of Bats in 2019 (MYLU)
```{r}
states_map <- map_data("state")

Regional_occupancy_trends_MYLU %>%
  filter(quantity == "psi_bar_t") %>%
  filter(year == "2019") %>%
  filter(region != c("Alberta", "British Columbia", "District of Columbia", "Manitoba", "New Brunswick", "Nova Scotia", "Ontario", "Prince Edward Island", "Quebec", "Saskatchewan", "Yukon")) %>% 
  mutate(region = str_to_lower(region)) %>%
  ggplot(aes(map_id = region, fill = mean)) + 
  geom_map(map = states_map, color = "black") + 
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map() + 
  scale_fill_viridis_c() + 
  labs(fill = "Mean probability", title = "Probability of occupancy for Little Brown bat by state in 2019")
```
However, in this map in 2019 we can see that the probably of seeing bats on the East Coast has decreased considerably. Even states midwest states that are more "newly" infected have felt the impact and have a decreased probably as well. 

#### Map of CHANGE of probability of Bats in 2019 (MYLU)
```{r}
states_map <- map_data("state")

Regional_occupancy_trends_MYLU %>%
  filter(quantity == "delta_bar_t") %>%
  filter(year == "2019") %>%
  filter(region != c("Alberta", "British Columbia", "Manitoba", "New Brunswick", "Nova Scotia", "Ontario", "Prince Edward Island", "Quebec", "Saskatchewan", "Yukon")) %>% 
  mutate(region = str_to_lower(region)) %>%
  ggplot(aes(map_id = region, fill = mean)) + 
  geom_map(map = states_map, color = "black") + 
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map() + 
  scale_fill_viridis_c() + 
  labs(fill = "Mean change in probability", title = "Change in probability of occupancy for Little Brown bat by state by 2019")
```
There are large decreases in probability in 2019. Areas recently hit such as the midwest and south have the strongest decline where as west seems either unaffected or only minorily so. The East Coast also has less dramatic decreases, likely because they had contact with WNS earlier in the decade. Those populations have either possibly developed a resilience or their population has been so decimated that little change can realisticlly be made. 

#### Statistical Significance 
```{r}
WNS_Occupancy_Binded_psi <- WNS_Occupancy_Binded %>%
  filter(quantity == "psi_bar_t")
```

```{r}
determination_psi.fit <- lm(mean ~ determination, data = WNS_Occupancy_Binded_psi) 
summary(determination_psi.fit) 
```
The No presence, Pd Positive, WNS positive and WNS supspect classifications are all statistically significant. However very few states had Pd Presumed or WNS Suspect classifications and the small sample could be what is making those variables significant. However, WNS Positive clearly shows a negative correlation with probability of seeing bats. This clearly shows that WNS is a cause of MYLU's decline.  

## Discussion: 
From my graphs, I can gather that the population of MYLU is slowly decreasing. While that rate might change at time periods, its overall trend does not. Furthermore, it is abundantly clear that WNS is spreading across the U.S. and relatively rapidly, covering twice the states it did at the beginning of the decade. Lastly, WNS has had a negative effect on the Little Brown Bat population and is the reason for their decline. 

While these conclusions are clear, their could be discrepancies with the real world. These were two unrelated data sets that I had to scale and wrangle in order to combine for comparison. In order to use the WNS data I had to assume that if a county within a state had WNS than the whole state had WNS, and if a state was identified as having WNS then it continued to have WNS for the rest of the study. This made the analysis less flexible and possibly less accurate. 

However, this could be improved upon by gathering proper data about the spread of WNS and rerunning this analysis. This could be bolstered my conclusion by monitoring at multiple bat sites to confirm my conclusions and further research by studying to see if bat populations rebound post initial infection. Research could even be undergone to see if some bats have built up a resilience to the fungus or if affect difference between bat species. Lastly, the effect of the bats decimation on insects and ecosystem services could be measured considering their importance to humans. 

## Works Cited: 

Clare, E. L., Barber, B. R., Sweeney B. W., Herbert P. D., & Fenton, M. B. (2011). Eating local: Influences of habitat on the Diet of Little Brown Bats (Myotis Lucifugus). Molecular Ecology, 20(8), 1772–1780. https://doi.org/10.1111/j.1365-294x.2011.05040.x 

Frick, W. F., Puechmaille, S. J., & Willis, C. K. R. (2016). White-Nose Syndrome in Bats. In Bats in the Anthropocene: Conservation of Bats in a Changing World (pp. 245–262). essay, Springer Nature.

Pettit, J. L., & O’Keefe, J. M. (2017). Impacts of white-nose syndrome observed during long-term monitoring of a midwestern bat community. Journal of Fish and Wildlife Management, 8(1), 69–78. https://doi.org/10.3996/102016-jfwm-077 

Reeder, D. M., Frank, C. L., Turner, G. G., Meteyer, C. U., Kurta, A., Britzke, E. R., Vodzak, M. E., Darling, S. R., Stihler, C. W., Hicks, A. C., Jacob, R., Grieneisen, L. E., Brownlee, S. A., Muller, L. K., & Blehert, D. S. (2012). Frequent arousal from hibernation linked to severity of infection and mortality in bats with white-nose syndrome. PLoS ONE, 7(6). https://doi.org/10.1371/journal.pone.0038920 

Udell, B.J., Straw, B.R, Cheng, T., Enns, K.D., Winfred, F., Gotthold, B.S, Irvine, K.M., Lausen, C., Loeb, S., Reichard, J., Rodhouse, T., Smith, D.A., Stratton, C., Thogmartin, W.E., Wiens, A.M., and Reichert, B.E., 2022, Status and Trends of North American Bats Summer Occupancy Analysis 2010-2019 Data Release: U.S. Geological Survey data release, https://doi.org/10.5066/P92JGACB.

U.S. Fish and Wildlife Service . (2015). Where is WNS now? . White Nose Syndrome Response Team. https://www.whitenosesyndrome.org/where-is-wns 
